\documentclass[12pt,a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{url}
\usepackage{microtype}
\usepackage{setspace}
\usepackage{titlesec}
\usepackage{abstract}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{float}
\usepackage{array}

\hypersetup{
  colorlinks=true,
  linkcolor=blue!60!black,
  citecolor=blue!60!black,
  urlcolor=blue!60!black
}

\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{gray!8},
  frame=single,
  rulecolor=\color{gray!40},
  breaklines=true,
  captionpos=b
}

\titleformat{\section}{\large\bfseries}{\thesection.}{0.5em}{}
\titleformat{\subsection}{\normalsize\bfseries}{\thesubsection.}{0.5em}{}

\onehalfspacing

% ---------------------------------------------------------------
\title{%
  \textbf{Lunar Ice Intelligence:}\\
  Physics-Informed Deep Learning for Water Ice Detection\\
  in Lunar Permanently Shadowed Regions
}

\author{
  Yago Almeida da Silva\\
  \small ORCID: \href{https://orcid.org/0009-0007-0094-0915}{0009-0007-0094-0915}\\
  \small \texttt{yagoalcontact@gmail.com}
}

\date{May 2026 \\ \small Preprint — not peer-reviewed}
% ---------------------------------------------------------------

\begin{document}

\maketitle

\begin{abstract}
We present \textit{Lunar Ice Intelligence}, an open-source platform for detecting
water ice in lunar permanently shadowed regions (PSRs) using physics-informed deep
learning. The system integrates real data from four NASA Lunar Reconnaissance Orbiter
(LRO) instruments — Diviner, Mini-RF, LROC WAC, and LAMP — with a multimodal neural
network (LunarCNN) that couples a convolutional image encoder with a physics-based
subsurface thermal encoder. Labels are derived exclusively from independent
instrument confirmations, avoiding the circularity common in PSR-based training sets.
A Double Deep Q-Network (DQN) agent performs autonomous rover navigation, maximising
ice probability while accounting for subsurface thermal state. On a benchmark of 14
catalogued locations drawn from four published datasets, the model achieves 86\%
accuracy (12/14), with all six negative controls correctly rejected and two lower-confidence
PSRs (Nobile, Amundsen) missed (F1 = 0.997, Recall = 1.000, val\_loss = 0.0109
on the validation set at 30 epochs). The platform
exposes a FastAPI backend and an interactive React web interface for point-and-click
ice probability analysis of the full 180$\times$360 lunar grid.
\end{abstract}

\vspace{0.5em}
\noindent\textbf{Keywords:} lunar water ice, permanently shadowed regions,
physics-informed neural network, deep reinforcement learning,
Lunar Reconnaissance Orbiter, subsurface thermal modelling, autonomous rover.

\hrule
\vspace{1em}

% ---------------------------------------------------------------
\section{Introduction}

The presence of water ice in lunar PSRs has been progressively confirmed by multiple
independent instruments over the past two decades. The LCROSS impact experiment
definitively detected H$_2$O in the Cabeus crater ejecta plume \citep{colaprete2010},
while Diviner temperature measurements identified cold traps capable of preserving
ice over geological timescales \citep{paige2010}. Complementary evidence was provided
by Mini-RF circular polarisation ratio (CPR) anomalies \citep{spudis2010} and LAMP
ultraviolet signatures \citep{gladstone2010}.

Despite this multi-instrument convergence, existing detection approaches typically
train models on PSR geometry alone, inadvertently reusing the same spatial prior used
to define the labels — a circularity that inflates reported performance. Furthermore,
few systems integrate subsurface thermal physics into the feature space, despite
surface temperature being a poor proxy for ice stability at depth.

This work addresses both limitations. We (1) construct labels using only
instrument-direct measurements with no overlap with the PSR spatial prior, (2) encode
subsurface temperature profiles derived from the Vasavada thermal diffusion model as
explicit input features, and (3) couple the ice detection model with a reinforcement
learning rover agent that uses subsurface state as part of its reward signal.

% ---------------------------------------------------------------
\section{Related Work}

\citet{mazarico2011} provided the foundational PSR catalogue from LOLA illumination
modelling. \citet{hayne2015} extended the Diviner analysis to a global hydration map.
\citet{williams2019} published the EPF dataset, providing direct surface-emission
temperature measurements inside PSRs. The thermal skin depth formulation used here
follows \citet{vasavada2012}, with thermal diffusivity $\kappa = 4.7 \times 10^{-7}$
m$^2$\,s$^{-1}$ appropriate for the uppermost lunar regolith.

Machine learning approaches to planetary surface analysis have been applied to Mars
and Moon imagery \citep{sato2014}, but physics-informed architectures integrating
multi-instrument LRO data with subsurface modelling have not, to our knowledge, been
previously published in open-source form.

% ---------------------------------------------------------------
\section{Data}

\subsection{Grid and Coordinate Convention}

All arrays use a global 180$\times$360 grid at 1$^\circ$/pixel. Grid indices
$(i, j)$ map to geographic coordinates as:
\begin{equation}
  \phi = i - 90, \qquad \lambda = j - 180
\end{equation}
where $\phi \in [-90^\circ, 90^\circ]$ is latitude and $\lambda \in [-180^\circ,
180^\circ]$ is longitude. This single coordinate convention is enforced across all
pipeline modules to prevent label misalignment.

\subsection{Instrument Datasets}

\begin{table}[H]
\centering
\caption{LRO instrument data integrated in the pipeline.}
\label{tab:data}
\begin{tabular}{lllr}
\toprule
Instrument & Data product & Acquisition method & Size \\
\midrule
Diviner LRO  & Polar temperature     & vsicurl GeoTIFF, $|\phi| \geq 60^\circ$ & $\sim$260 KB\\
Diviner EPF  & PSR direct temps      & PDS direct download       & $\sim$15 MB \\
LROC WAC     & 32$\times$32 patches  & vsicurl USGS COG per PSR  & $\sim$1 MB  \\
Mini-RF      & CPR polar radar       & HTTP range requests       & 59 MB       \\
LAMP         & UV H$_2$O signature   & PDS FITS download         & $\sim$10 MB \\
\bottomrule
\end{tabular}
\end{table}

Selective streaming avoids downloading full-resolution global mosaics
(typically $>$10 GB). Only polar strips ($|\phi| \geq 60^\circ$) are retrieved
from Diviner; LROC patches are fetched per-PSR via Cloud-Optimised GeoTIFF
(COG) range requests against the USGS tile server.

\subsection{Label Construction}

Labels are binary ($y \in \{0, 1\}$) and are derived from three independent
instrument sources fused with per-source confidence weights:

\begin{table}[H]
\centering
\caption{Label sources, confidence weights, and positive counts.}
\label{tab:labels}
\begin{tabular}{llrl}
\toprule
Source & Confidence & Positives & Reference \\
\midrule
LCROSS Cabeus impact    & 1.0 & 1      & \citet{colaprete2010} \\
Diviner EPF direct      & 1.0 & $\sim$6 & \citet{williams2019} \\
LAMP UV Shackleton      & 0.9 & 2      & \citet{gladstone2010} \\
Mini-RF CPR $> 1.0$     & 0.8 & 14{,}655 & \citet{spudis2010}  \\
Diviner cold trap       & 0.7 & $\sim$50 & \citet{paige2010}  \\
\midrule
\textbf{Total}          &     & \textbf{14{,}656} & \\
\bottomrule
\end{tabular}
\end{table}

PSR geometry is used only to locate known confirmed sites, never as a positive
label itself. This decouples the training signal from the spatial prior and
prevents performance inflation through circularity.

% ---------------------------------------------------------------
\section{Model Architecture}

\subsection{LunarCNN}

The detection model is a three-branch multimodal network (Listing~\ref{lst:arch}):

\paragraph{CNNEncoder.} Processes 64$\times$64 greyscale image patches through
three Conv–BN–ReLU–MaxPool blocks (channels: 1$\to$16$\to$32$\to$64),
followed by a fully-connected layer (4096$\to$128) producing a 128-dimensional feature vector.

\paragraph{PhysicsEncoder.} Receives a 5-dimensional physics feature vector:
\begin{equation}
  \mathbf{f}_p = \bigl[S_\text{norm},\ \phi_\text{norm},\ T_{0.1},\ T_{0.5},\ T_{1.0}\bigr]
\end{equation}
where $S_\text{norm} = S/1361$ is insolation normalised to the solar constant,
$\phi_\text{norm} = \phi/90$ is normalised latitude, and $T_{0.1}$, $T_{0.5}$,
$T_{1.0}$ are subsurface temperatures at 0.1, 0.5, and 1.0 m depth (normalised).
A three-layer MLP (5$\to$32$\to$64$\to$64, BatchNorm + ReLU) produces a
64-dimensional output.

\paragraph{FusionHead.} Concatenates the CNN (128-D) and physics (64-D) vectors
into a 192-D representation, then applies a linear stack
$192 \to 128 \to 64 \to 1$ with sigmoid activation to produce $\hat{p} = P(\text{ice})$.

\begin{lstlisting}[language=Python, caption={LunarCNN forward pass (simplified).}, label=lst:arch]
img_feat   = self.cnn_encoder(image)        # (B, 128)
phys_feat  = self.physics_encoder(physics)  # (B, 64)
fused      = torch.cat([img_feat, phys_feat], dim=1)  # (B, 192)
p_ice      = self.fusion_head(fused)        # (B, 1)
\end{lstlisting}

\subsection{Subsurface Thermal Model}

Subsurface temperature profiles follow the analytical solution to the one-dimensional
thermal diffusion equation for a periodically forced surface \citep{vasavada2012}:
\begin{equation}
  T(z) = \bar{T} + \left(T_s - \bar{T}\right) \exp\!\left(-\frac{z}{z_\text{skin}}\right)
\end{equation}
where $T_s$ is the instantaneous surface temperature, $\bar{T}$ is the mean annual
temperature, and the thermal skin depth is:
\begin{equation}
  z_\text{skin} = \sqrt{\frac{\kappa P}{\pi}} \approx 0.62\ \text{m}
\end{equation}
with $\kappa = 4.7 \times 10^{-7}$ m$^2$\,s$^{-1}$ and $P = 2.551 \times 10^6$ s
(lunar synodic period). Surface temperature is derived from radiative equilibrium:
\begin{equation}
  T_s = \left(\frac{(1 - A)\,S\cos\theta + \epsilon\sigma T_\infty^4}{\epsilon\sigma}\right)^{1/4}
\end{equation}
with albedo $A = 0.12$, emissivity $\epsilon = 0.95$, and $T_\infty = 3$ K.
Profiles are computed at 20 depth layers spanning 0–2 m and stored as
\texttt{temperatura\_subsolo.npy} (shape $180 \times 360 \times 20$).

\subsection{Training}

\begin{itemize}
  \item \textbf{Dataset:} 58{,}624 examples; 14{,}656 positives (25\%)
  \item \textbf{Loss:} Weighted binary cross-entropy (handles class imbalance)
  \item \textbf{Optimiser:} Adam, lr = $10^{-3}$
  \item \textbf{Schedule:} CosineAnnealingLR
  \item \textbf{Epochs:} 30 (configurable via \texttt{TRAIN\_EPOCHS})
  \item \textbf{Uncertainty:} Monte Carlo Dropout (30 forward passes) for variance estimation
\end{itemize}

% ---------------------------------------------------------------
\section{Autonomous Rover Navigation}

\subsection{Markov Decision Process Formulation}

The rover navigation task is cast as an episodic MDP with:

\paragraph{State space ($\text{obs\_dim} = 6$):}
\begin{equation}
  \mathbf{s} = \bigl[\phi_n,\ \lambda_n,\ E_n,\ S_n,\ \hat{p}_\text{ice},\ T^{(0.5\text{m})}_n\bigr]
\end{equation}
where $\phi_n$, $\lambda_n$ are normalised grid coordinates, $E_n$ is normalised
remaining energy, $S_n$ is normalised insolation, $\hat{p}_\text{ice}$ is the CNN
ice probability, and $T^{(0.5\text{m})}_n$ is the normalised 0.5 m subsurface
temperature at the current position.

\paragraph{Action space:} Four discrete moves — North, South, East, West.

\paragraph{Reward:}
\begin{equation}
  r = \Delta\hat{p}_\text{ice} - c_E\,\Delta E + r_\text{discover} + r_\text{subsurface}
\end{equation}
\begin{equation}
  r_\text{subsurface} = \max\!\bigl(0,\ 1 - T^{(0.5\text{m})}_n\bigr) \times 0.4
\end{equation}
The subsurface bonus incentivises the agent to prefer cold-at-depth regions even
when surface temperature is ambiguous — a physically motivated exploration prior.

\subsection{Double DQN Agent}

The agent uses Double Deep Q-Learning \citep{vanHasselt2016} with:
\begin{itemize}
  \item Experience replay buffer of 10{,}000 transitions
  \item Target network with soft update: $\theta^- \leftarrow \tau\theta + (1-\tau)\theta^-$, $\tau = 0.005$
  \item $\varepsilon$-greedy exploration with exponential decay over 5{,}000 steps
  \item Q-network: MLP $6 \to 128 \to 128 \to 4$
  \item Training: 500 episodes
\end{itemize}

% ---------------------------------------------------------------
\section{Results}

\subsection{Detection Performance}

\begin{table}[H]
\centering
\caption{CNN training and validation metrics (30 epochs, CPU).}
\label{tab:results}
\begin{tabular}{lr}
\toprule
Metric & Value \\
\midrule
val\_loss & 0.0109 \\
F1 score  & 0.997  \\
Recall    & 1.000  \\
Accuracy  & 0.998  \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Benchmark Against Published Catalogues}

\begin{table}[H]
\centering
\caption{Benchmark on 14 catalogued locations from four publications.
  \checkmark~= correct prediction; $\times$~= incorrect.}
\label{tab:benchmark}
\begin{tabular}{llccc}
\toprule
Location & Source & $\hat{p}$ & Expected & Result \\
\midrule
\multicolumn{5}{l}{\textit{South pole — confirmed ice}} \\
Cabeus      & \citet{colaprete2010} & 0.983 & ICE & \checkmark \\
Shackleton  & \citet{gladstone2010} & 1.000 & ICE & \checkmark \\
Haworth     & \citet{spudis2010}    & 1.000 & ICE & \checkmark \\
Nobile      & \citet{hayne2015}     & 0.000 & ICE & $\times$   \\
Amundsen    & \citet{mazarico2011}  & 0.000 & ICE & $\times$   \\
\midrule
\multicolumn{5}{l}{\textit{North pole — confirmed ice}} \\
Hermite     & \citet{paige2010}     & 1.000 & ICE & \checkmark \\
Peary       & \citet{spudis2010}    & 1.000 & ICE & \checkmark \\
Whipple     & \citet{mazarico2011}  & 1.000 & ICE & \checkmark \\
\midrule
\multicolumn{5}{l}{\textit{Negative controls — no ice}} \\
Equator 0°N & —                     & 0.000 & NO ICE & \checkmark \\
Mare Tranq. & \citet{sato2014}      & 0.000 & NO ICE & \checkmark \\
Copernicus  & —                     & 0.000 & NO ICE & \checkmark \\
Tycho       & —                     & 0.000 & NO ICE & \checkmark \\
Mid-lat 40°N & —                    & 0.000 & NO ICE & \checkmark \\
Mid-lat 40°S & —                    & 0.000 & NO ICE & \checkmark \\
\midrule
\textbf{Total} & & & & \textbf{12/14 (86\%)} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Scientific Validation}

Validation against 6 PSRs with known Diviner EPF measurements confirms that
predicted probabilities are consistent with published surface temperatures and
cold-trap stability criteria ($T < 110$ K, \citealt{paige2010}).

The two benchmark failures (Nobile, $\hat{p} = 0.000$; Amundsen, $\hat{p} = 0.000$)
are both lower-confidence PSRs (conf~$\leq 0.75$) located at non-extreme south-polar
longitudes ($\lambda = +53.5^\circ$ and $+84.7^\circ$). Neither site has direct
LCROSS, EPF, or LAMP confirmation; both rely on Diviner cold-trap modelling
\citep{hayne2015} and LOLA illumination \citep{mazarico2011} alone. Their absence
from the Mini-RF CPR label set (the dominant positive source at $\geq 60^\circ$ S)
means the model has insufficient positive signal to generalise to these mid-longitude
polar targets — a direct consequence of the Mini-RF polar strip coverage pattern.

% ---------------------------------------------------------------
\section{System Architecture}

The platform consists of three layers:

\paragraph{Backend (FastAPI).} Serves five endpoints with rate limiting (slowapi).
The \texttt{/analisar} endpoint returns eight fields per query:
\texttt{probabilidade\_gelo}, \texttt{variancia}, \texttt{confianca},
\texttt{temperatura}, \texttt{temperatura\_subsolo[3]}, \texttt{insolacao},
\texttt{insolacao\_atual}, \texttt{fase\_lunar}. Authentication uses
\texttt{X-API-Key} headers with CSP and HSTS headers in production.

\paragraph{Frontend (React + Leaflet).} A landing page with eight sections —
including an interactive lunar map where the user clicks to trigger ice analysis —
served via Vite at port 5173. Rover trajectories are drawn in real time as
grid-to-degree converted polylines.

\paragraph{Infrastructure.} Docker multi-stage build on \texttt{python:3.13-slim};
GitHub Actions CI; data and model volumes persist across container rebuilds.

% ---------------------------------------------------------------
\section{Discussion}

The 86\% benchmark accuracy (12/14) reflects a genuine model limitation rather than
a calibration artefact. All six negative controls and six of eight PSRs are correctly
classified; the two failures (Nobile, Amundsen) are the lowest-confidence PSRs in
the catalogue (conf~$\leq 0.75$) and lack direct instrument confirmation beyond
Diviner cold-trap modelling. The value of the system lies not in the aggregate
benchmark figure but in (1) the absence of circular label construction, (2) the
explicit physics encoding of subsurface temperature, and (3) the provision of
per-prediction uncertainty via MC Dropout.

The subsurface thermal bonus in the reward function of the DQN agent constitutes
a physically motivated exploration prior. Regions cold at 0.5 m depth are
preferentially visited even when surface albedo or insolation measurements are
ambiguous — a heuristic directly grounded in ice stability theory \citep{paige2010}.

A limitation of the current label set is the dominance of Mini-RF CPR positives
(99.99\% of positives), whose confidence weight (0.8) is lower than direct
EPF measurements. Future work should prioritise integrating additional Diviner EPF
observations and LAMP count-rate maps to strengthen the high-confidence label pool.

% ---------------------------------------------------------------
\section{Conclusion}

We have described Lunar Ice Intelligence, a physics-informed deep learning platform
for lunar water ice detection. Key contributions are: a multimodal LunarCNN
integrating image and subsurface thermal features; a label pipeline free of
PSR-geometry circularity; a Double DQN rover agent with a physics-grounded
subsurface reward; and an end-to-end open-source system from raw LRO data to
interactive web interface. The software is released under the Apache License 2.0
and archived with a CITATION.cff for reproducible citation.

% ---------------------------------------------------------------
\section*{Software Availability}

Source code, trained weights (\texttt{pesos.pth}, \texttt{rl\_pesos.pth}),
and data pipeline are available under the Apache License 2.0.
Cite as: Almeida da Silva, Y. (2026). \textit{Lunar Ice Intelligence}.
ORCID: \href{https://orcid.org/0009-0007-0094-0915}{0009-0007-0094-0915}.

% ---------------------------------------------------------------
\section*{Acknowledgements}

The author acknowledges NASA's Planetary Data System (PDS) and the LRO science
teams (Diviner, Mini-RF, LROC, LAMP) for making the datasets publicly available.

% ---------------------------------------------------------------
\begin{thebibliography}{99}

\bibitem[Colaprete et al.(2010)]{colaprete2010}
Colaprete, A., et al. (2010).
Detection of Water in the LCROSS Ejecta Plume.
\textit{Science}, 330, 463--468.
\url{https://doi.org/10.1126/science.1186986}

\bibitem[Gladstone et al.(2010)]{gladstone2010}
Gladstone, G. R., et al. (2010).
LAMP: The Lyman Alpha Mapping Project on the Lunar Reconnaissance Orbiter Mission.
\textit{Science}, 330, 472--476.
\url{https://doi.org/10.1126/science.1197874}

\bibitem[Hayne et al.(2015)]{hayne2015}
Hayne, P. O., et al. (2015).
Evidence for Exposed Water Ice in the Moon's South Polar Regions from Lunar
Reconnaissance Orbiter Ultraviolet Albedo and Temperature Measurements.
\textit{Journal of Geophysical Research: Planets}, 120, 1599--1615.
\url{https://doi.org/10.1002/2014JE004797}

\bibitem[Mazarico et al.(2011)]{mazarico2011}
Mazarico, E., et al. (2011).
Illumination conditions of the lunar polar regions using LOLA topography.
\textit{Icarus}, 211, 1066--1081.
\url{https://doi.org/10.1016/j.icarus.2010.10.030}

\bibitem[Paige et al.(2010)]{paige2010}
Paige, D. A., et al. (2010).
Diviner Lunar Radiometer Observations of Cold Traps in the Moon's South Polar Region.
\textit{Science}, 330, 479--482.
\url{https://doi.org/10.1126/science.1187726}

\bibitem[Sato et al.(2014)]{sato2014}
Sato, H., et al. (2014).
Resolved Hapke parameter maps of the Moon.
\textit{Journal of Geophysical Research: Planets}, 119, 1775--1805.
\url{https://doi.org/10.1002/2013JE004580}

\bibitem[Spudis et al.(2010)]{spudis2010}
Spudis, P. D., et al. (2010).
Initial results for the north pole of the Moon from Mini-RF, the Miniature Radio
Frequency Radar.
\textit{Geophysical Research Letters}, 37, L06204.
\url{https://doi.org/10.1029/2009GL042259}

\bibitem[van Hasselt et al.(2016)]{vanHasselt2016}
van Hasselt, H., Guez, A., \& Silver, D. (2016).
Deep Reinforcement Learning with Double Q-learning.
\textit{Proceedings of the AAAI Conference on Artificial Intelligence}, 30(1).

\bibitem[Vasavada et al.(2012)]{vasavada2012}
Vasavada, A. R., et al. (2012).
Lunar equatorial surface temperatures and regolith properties from the Diviner
Lunar Radiometer Experiment.
\textit{Journal of Geophysical Research: Planets}, 117, E00H18.
\url{https://doi.org/10.1029/2011JE003987}

\bibitem[Williams et al.(2019)]{williams2019}
Williams, J.-P., et al. (2019).
Diviner Lunar Radiometer Experiment eight-year special observation data record:
Reduction in data gaps and improvements in calibration, spatial sampling, and
noise.
\textit{Journal of Geophysical Research: Planets}, 124, 2078--2088.
\url{https://doi.org/10.1029/2018JE005803}

\end{thebibliography}

\end{document}
